\thetitle

### 2020-07-01 writing/challenges.tex
split introduction

\section{Does Blame Matter?}

A spectre is haunting gradual typing --- the spectre of blame.  All works
on gradual typing incantate spells either to incarnate blame as they see
fit or to avoid facing it.  But unfortunately, despite all the effort that
has gone to trap blame with models and theorems, its soul remains elusive.

Take for instance the so-called blame theorem~\cite{mf-toplas-2009,
tf-dls-2006, wf-esop-2009}. It reveals blame as an essential indicator of
gradual type soundness; sound gradual type systems never blame typed code.
On one hand, this helps language designers benefit from blame to
(partially) validate their gradual type systems~\cite{gf-icfp-2018}. On
the other hand though, it reduces blame to a proof apparatus.  Worse,
projecting the theorem to a lesson for the programmer leads to misleading
conclusions about where the bug is. The problem  may not be with the
blamed untyped code but with one of the annotations of the typed code.
Similarly, the recent complete monitoring framework~\cite{gfd-oopsla-2019}
sheds limited light to the pragmatic nature of blame for gradual types. It
shows that blame captures some aspects of the provenance of the value that
fails a runtime type check, but again this explanation falls short of
tangible information for the programmer. In effect programmers are left
without any guidance about how to turn information from blame to the
location of bugs.  And this current lack of a principled connection
between blame and bugs ---  the raison d'\^etre of blame --- puts in
question the value of blame for gradual types.

This paper aims to answer a single question: \emph{Does blame add value to
gradual typing?} We posit that blame does add value to gradual typing only
if it conveys to programmers extra information that is useful for
debugging. Specifically, for blame to be valuable there needs to be a
systematic relation between blame and the location of bugs.

To examine the truth of this condition, we follow recent work on the
pragmatics of blame for contracts~\cite{lksfd-popl-2020}. First, we stray
from the usual theoretical treatment of blame in the gradual typing
literature and we investigate the value of blame with an empirical
methodology. Its empirical nature gives us the opportunity to investigate how blame copes
against actual bugs in real implementations of gradual type systems.

Second, the central piece of the methodology is a connection between blame
and bugs inspired straight from the experience of programming with gradual
types.  When facing a run time type error in a mix-typed program, programmers often try to add
type annotations to their programs hoping to turn the  run time type error
into a static one that reveals where the code and its types clash.  In
this setting, blame can suggest what part of the program to migrate from
untyped to typed next.  Therefore, if heeding blame culminates to a static type
error, blame and the location of the bug are directly connected.  Based on
this observation, our methodology explores the migration of mix-typed programs with
run time type errors to validate whether blame guides their migration to an
ill-typed  version that pinpoints the location of the
bug.

>>>>

\section{Challenges}

Creating a method for comparing blame assignment strategies poses
several severe challenges.

### 2020-07-01 writing/introduction.tex
split introduction

A spectre is haunting gradual typing --- the spectre of blame.  All works
on gradual typing incantate spells either to incarnate blame as they see
fit or to avoid facing it.  But unfortunately, despite all the effort that
has gone to trap blame with models and theorems, its soul remains elusive.

Take for instance the so-called blame theorem~\cite{mf-toplas-2009,
tf-dls-2006, wf-esop-2009}. It reveals blame as an essential indicator of
gradual type soundness; sound gradual type systems never blame typed code.
On one hand, this helps language designers benefit from blame to
(partially) validate their gradual type systems~\cite{gf-icfp-2018}. On
the other hand though, it reduces blame to a proof apparatus.  Worse,
projecting the theorem to a lesson for the programmer leads to misleading
conclusions about where the bug is. The problem  may not be with the
blamed untyped code but with one of the annotations of the typed code.
Similarly, the recent complete monitoring framework~\cite{gfd-oopsla-2019}
sheds limited light to the pragmatic nature of blame for gradual types. It
shows that blame captures some aspects of the provenance of the value that
fails a runtime type check, but again this explanation falls short of
tangible information for the programmer. In effect programmers are left
without any guidance about how to turn information from blame to the
location of bugs.  And this current lack of a principled connection
between blame and bugs ---  the raison d'\^etre of blame --- puts in
question the value of blame for gradual types.

This paper aims to answer a single question: \emph{Does blame add value to
gradual typing?} We posit that blame does add value to gradual typing only
if it conveys to programmers extra information that is useful for
debugging. Specifically, for blame to be valuable there needs to be a
systematic relation between blame and the location of bugs.

To examine the truth of this condition, we follow recent work on the
pragmatics of blame for contracts~\cite{lksfd-popl-2020}. First, we stray
from the usual theoretical treatment of blame in the gradual typing
literature and we investigate the value of blame with an empirical
methodology. Its empirical nature gives us the opportunity to investigate how blame copes
against actual bugs in real implementations of gradual type systems.

Second, the central piece of the methodology is a connection between blame
and bugs inspired straight from the experience of programming with gradual
types.  When facing a run time type error in a mix-typed program, programmers often try to add
type annotations to their programs hoping to turn the  run time type error
into a static one that reveals where the code and its types clash.  In
this setting, blame can suggest what part of the program to migrate from
untyped to typed next.  Therefore, if heeding blame culminates to a static type
error, blame and the location of the bug are directly connected.  Based on
this observation, our methodology explores the migration of mix-typed programs with
run time type errors to validate whether blame guides their migration to an
ill-typed  version that pinpoints the location of the
bug.

Realizing these ideas into a robust experiment requires surpassing three challenges.  First,
gradual typing is a heterogeneous landscape. Herein, we focus on three
points of the landscape: the natural approach of Typed
Racket~\cite{tf-dls-2006,tf-popl-2008,tfffgksst-snapl-2017,tf-icfp-2010},
the transient one of Reticulated
Python~\cite{vsc-dls-2019,vss-popl-2017,vksb-dls-2014} and  the optional
gradual typing of the likes of Typescript. Each of these points comes with
a distinct blaming strategy and together they represent all proposed
strategies so far. In addition, instead of experimenting with the three
approaches in isolation, we have implemented transient and optional
gradual typing as variants of  Typed Racket to seize the opportunity to
compare them.  However, implementations on top of the same platform are
not enough for an apples-to-apples comparison. The three approaches differ
not only on their blaming strategies but also on the type checks they
perform. To untangle the two, we supplement our experiment with two extra
modes. The \emph{vanilla} mode  ignores blame and instead uses information
from dynamic type error messages to select how to proceed with migration. Hence it
separates the effect of blame on the experiment from that of the type
checks. The \emph{control} mode randomly selects which part of the program to migrate
and serves as the null hypothesis control for the overall methodology.

The second challenge stems from the lack of a sufficiently large available
corpus of type errors for a meaningful experiment. Similar to
~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in
typed programs  with mutation analysis~\cite{lipton1971fault,
demillo1978hints, jia2011analysis}. This results into a large 
number of buggy programs whose mix-typed versions are suitable subjects
for our experiment. However, mutation analysis
traditionally  aims to inject bugs that are sophisticated enough to
challenge extensive test suites. As a consequence, the ill-typed mutants we care about here are
usually deemed \emph{incompetent} and mutation analysis frameworks are
fine-tuned to avoid them. To overcome this problem, we design a
set of mutators specific for type errors inspired by common errors when
programming with in Typed Racket.  These type errors cover
a wide range of Typed Racket's type system and moreover cause reliably runtime
type errors  in mix-typed versions of mutants.

The final challenge deals with the fact that bugs in mix-typed programs
occur not only in the code but also in type
annotations~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}. Thus
our experiment needs to also explore the relation between blame and the
location of bugs in case of wrong type annotations. Similarly to the previous challenge, we
deal with this one by designing a set of mutators that ``break'' type
annotations but leave programs otherwise intact.

To sum up, the maim {\bf contribution} of this paper is an empirical  methodology for
evaluating blame for gradual type systems. The methodology applies across the
landscape of gradual typing and allows to compare its different points. 
  As a result, this paper also contributes the first comparative
empirical study of natural, transient and optional gradual typing
with respect to blame. [A couple of sentences that summarize the results
her.]

The remainder of the paper is organized as follows.
Section~\ref{sec:landscape} revisits the landscape of gradual typing from
the perspective of the different checking and blame strategies.
Section~\ref{sec:ideas} presents in detail the main ideas behind our
methodology and section~\ref{sec:experiment} explains the specifics of our
experiment. Section~\ref{sec:results} gathers our results and
section~\ref{sec:discussion} discusses their implications. The paper
concludes with a survey of related work in section~\ref{sec:related} and 
a few final remarks in section~\ref{sec:conclusion}.

>>>>

Since its very beginning [cite STH], the theory of gradual typing
comes with a blame theorem. ``Well-typed [components]\footnote{The
original authors got this word wrong, and more.} can't go wrong'' has
become the slogan. The idea is that blame assignment can help the
practicing programmers find impedance mismatches between the types
they added and the remaining untyped components.

Practical gradual type systems ignore blame completely. Industrial
systems [lots of citations. footnotes] exploit types for assistance
with IDE actions and for finding typos in code. Then they remove them
and rely on the built-in safety checks of the underlying programming
language to catch problems.

This contrast raises the question whether

     blame adds value to gradual typing
     and especially the working programmer.

To obtain an answer, we have developed a method (not methodology!)  to
evaluate blame assignment systems from the gradual typing
world. Roughly speaking, the method simulates a programmer who follows
the above-cited Wadler-Findler slogan. When the run-time system
signals a violation, the programmer conjectures that the addition of 
types to untyped components turns the dynamic violation into a static
type error and thus reveals the impedance mismatch.

The paper presents the evaluation method and the results of applying
the method to three distinct blame assignment systems.

- explain the systems
- explain the method
- preview the results

- weave roadmap into these paragraphs
